---
title: "hhh4addon: extending the functionality of surveillance:hhh4"
author: "Johannes Bracher"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Purpose of the `hhh4addon` package

The `hhh4addon` package modifies and extends  the functionality of the `surveillance` package, more specifically the model class implemented in the function `hhh4`. **Note that `hhh4addon` is currently only available in a development version.** It adds the following features:

- Fitting model versions with distributed-lags.
- Computation of predictive and stationary first and second moments.

As `hhh4addon` can only be used in combination with `surveillance` we assume some basic familiarity with this package in the following.


## The `hhh4` model class and its extension to distributed lags

We only give a very brief description of the `hhh4` modelling framework, details can be found for instance in `vignette("hhh4")` and `vignette("hhh4_spacetime")` from the `surveillance` package. Counts $X_{it}$ in unit $i$ at time $t$ are modelled as

$$
X_{it} \mid \mathbf{X}_{t - 1} \sim \text{NegBin}(\mu_{it}, \psi_i); X_{it} \bot X_{jt} \mid \mathbf{X}_{t - 1}\\
\mu_{it} = e_{it}\nu_{it} + \lambda_{it}X_{i, t - 1} + \phi_{it}\sum_{j \neq i} w_{ji}X_{j, t - 1}
$$
where the negative binomial distribution is parametrized by its mean $\mu_{it}$ and an overdispersion perameter $\psi_i$ such that the conditional variance $\text{Var}(X_t \mid X_{t - 1}) = \mu_{it}\cdot(1 + \psi_i\mu_{it})$. While $e_{it}$ is a population offset, the remaining parameters of the mean structure are modelled in a log-linear fashion. While in principle covariates can enter here, it is common to let only an intercept and sine/cosine terms for seasonality enter:
$$
\log(\nu_{it}) = \alpha^{(\nu)}_i + \beta^{(\nu)}_i \sin(2\pi t/\omega) + \gamma^{(\nu)}_i \cos(2\pi t/\omega)\\
\log(\lambda_{it}) = \alpha^{(\lambda)}_i + \beta^{(\lambda)}_i \sin(2\pi t/\omega) + \gamma^{(\lambda)}_i \cos(2\pi t/\omega)\\
\log(\phi_{it}) = \alpha^{(\phi)}_i + \beta^{(\phi)}_i \sin(2\pi t/\omega) + \gamma^{(\phi)}_i \cos(2\pi t/\omega)\\
$$
Whether unit-specific parameters are necessary and identifiable depends on the data at hand, see the more general vignettes from `surveillance` for more information on model building.

The added functionality which `hhh4addon` offers is to allow for distributed lags, i.e. models with a conditional mean of the form
$$
\mu_{it} = \nu_{it} + \lambda_{it}\sum_{d = 1}^D \lfloor u_d\rfloor X_{t - d} + \phi_{it}\sum_{j\neq i}\sum_{d = 1}^D w_{ji}\lfloor u_d\rfloor X_{j, t - d}
$$

where the weights $u_d$ are normalized such that $\sum_{d = 1}^D \lfloor u_d\rfloor = 1$ before they enter. This means that instead of the previous observation $X_{t - 1}$ a weighted average of the $D$ preceding observations $X_{t - D},\dots, X_{t - 1}$ enters. Currently only a geometric, i.e. exponentially decaying lag structure is implemented:
$$
u_d = p(1 - p)^{d - 1}
$$
We exemplify this with an with a simple univariate analysis of the `salmonella` data from `surveillance`. All syntax also translates to the multivariate case.
```{r, fig.show='hold'}
library(surveillance, quietly = TRUE)
library(hhh4addon)
data("salmonella.agona")
## convert old "disProg" to new "sts" data class
salmonella <- disProg2sts(salmonella.agona)

plot(salmonella)
```

First we fit a regular `hhh4` model with seasonality in both components:
```{r, message=FALSE, warning=FALSE}
control_salmonella <- list(end = list(f = addSeason2formula(~ 1)),
                            ar = list(f = addSeason2formula(~ 1), lag = 1),
                            family = "NegBinM", subset = 6:312)
fit_salmonella <- hhh4_lag(salmonella, control_salmonella)
AIC(fit_salmonella)
```

When `hhh4addon` is loaded, a modified version `hhh4lag` of `hhh4` is available which allows for some additional specifications in the `control` list. Specifically the following arguments can be specified in the `ar` (and potentially `ne`) components:

- `use_distributed_lag`: de-activates the usual `lag` argument and activates the use of the three following arguments
- `funct_lag`: a function to compute distributed lags from simple first lags given some parameter value and a maximum number of lags. Currently only the default function `geometric_lag` is implemented.
- `par_lag`: the parameter entering into `funct_lag`; for `geometric_lag` this is $p$ (which corresponds to the un-normalized weight of the first lag)
- `max_lag`: the maximum number of lags entering into `funct_lag`

The folowing code allows to fit a model with a geometric lag structure, a fixed value of $p = 0.8$ and five lags. Note that for such a model we can only use the data from week 6 onwards for fitting (the lagged observations are not available for weeks 1 through 5).

```{r}
control_salmonella.08 <- list(end = list(f = addSeason2formula(~ 1)),
                            ar = list(f = addSeason2formula(~ 1), use_distr_lag = TRUE, 
                                      par_lag = 0.8, max_lag = 5),
                            family = "NegBinM", subset = 6:312)
fit_salmonella.08 <- hhh4_lag(salmonella, control_salmonella.08)
plot(fit_salmonella.08)
AIC(fit_salmonella.08) # 1225.157
```

We can see that this model is better than the above model with first lags. To estimate $p$ from the data we can use wrapper `fit_par_lag` which evaluates `hhh4_lag` for a set of values for `par_lag` provided in an argument `range_par`:

```{r}
control_salmonella.flex <- list(end = list(f = addSeason2formula(~ 1)),
                            ar = list(f = addSeason2formula(~ 1), use_distr_lag = TRUE, 
                                      max_lag = 5),
                            family = "NegBinM", subset = 6:312)
range_par <- seq(0.1, 1, by = 0.01)
fit_salmonella.flex <- fit_par_lag(salmonella, control_salmonella.flex, range_par = range_par)
```
The return object of this function contains two objects: the `hhh4` object from the fit with the lowest AIC (`best_mod`) and a vector giving the AIC values for models using the different values of par_lag.
```{r}
plot(fit_salmonella.flex$best_mod)
summary(fit_salmonella.flex$best_mod)
plot(range_par, fit_salmonella.flex$AICs, type = "l", xlab = "p", ylab = "AIC")
min(fit_salmonella.flex$AICs)
```
The best fit is attained with a model using $p = 0.56$, i.e. almost half of the contribution of the epidemic contribution comes from lags of order larger than one.

A remark on the computation of the AICs: The AIC of a model which was fitted with `fit_par_lag` is 2 points higher than that of a model with the same value of `par_lag`, but specified manually instead of being estimated. This is due to the loss of one degree of freedom.
```{r}
control_salmonella.056 <- list(end = list(f = addSeason2formula(~ 1)),
                            ar = list(f = addSeason2formula(~ 1), use_distr_lag = TRUE, 
                                      par_lag = 0.56, max_lag = 5),
                            family = "NegBinM", subset = 6:312)
fit_salmonella.056 <- hhh4_lag(salmonella, control_salmonella.056)
AIC(fit_salmonella.056)
AIC(fit_salmonella.flex$best_mod)
```

## Predictive and periodically stationary moments

The second and more mature functionality of `hhh4addon` concerns predictive and periodically stationary moments. These quantities can be computed without the need for simulation for the `hhh4` class, see Held, Meyer and Bracher (2017) and Bracher and Held (2017) for the theoretical background.

To demonstrate this we re-fit the above models on a a subset of the `salmonella` data and keep the rest for validation of our predictions.
```{r}
control_salmonella.sub <- list(end = list(f = addSeason2formula(~ 1), lag = 1),
                            ar = list(f = addSeason2formula(~ 1), lag = 1),
                            family = "NegBinM", subset = 6:260)
fit_salmonella.sub <- hhh4(salmonella, control_salmonella.sub)
```
Predictive moments can now be calculated and plotted as follows:
```{r}
pred_mom <- predictive_moments(fit_salmonella.sub, t_condition = 260, lgt = 52, return_Sigma = TRUE)
plot(fit_salmonella.sub)
fanplot_prediction(pred_mom, add = TRUE)
fanplot_prediction(pred_mom, add = FALSE)
```

The fanplots here are based on negative binomial approximations of the predictive distributions via the first two moments. We can also use these predictive moments to evaluate the Dawid-Sebastiani score, a proper scoring rule for predictive model assessment:
```{r}
ds_score_hhh4(pred_mom)
```

We use this to compare a model with one lag to a model with geometric lags.
```{r}
# fit the model with geometric lags
control_salmonella.sub.flex <- list(end = list(f = addSeason2formula(~ 1)),
                            ar = list(f = addSeason2formula(~ 1), use_distr_lag = TRUE, 
                                      par_lag = 0.8, max_lag = 5),
                            family = "NegBinM", subset = 6:260)
range_par <- seq(0.1, 1, by = 0.01)
fit_salmonella.sub.flex <- fit_par_lag(salmonella, control_salmonella.sub.flex, range_par = range_par)
# get prediction:
pred_mom.flex <- predictive_moments(fit_salmonella.sub.flex$best_mod, t_condition = 260, lgt = 52, return_Sigma = TRUE)
ds_score_hhh4(pred_mom.flex)
```
Like other predictive scores the David-Sebastiani score is negatively oriented, i.e. the prediction from the model `fit_salmonella.sub.flex` with geometric lags is slightly better than the one from the simpler `fit_salmonella.sub`.
